Follow-up Samples

### 2017-02-21 submission/manuscript_outline_20161024.Rmd

**Results:** The bacterial microbiome in pre and post lesion removal samples for the adenoma group were more similar to each other than the carcinoma group (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between the pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A model built to classify model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.  Follow up samples had a decrease in the positive probability (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  This decrease in positive probability was also observed for the initial sample model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g. *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).

**Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post lesion removal samples.  With individuals with carcinoma having more drastic differences to the overall community then those with adenoma.  Changes to commensal bacteria were some of the most important variables for model classification, suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].

Recently, there has been promising work on the bacterial microbiome and it's ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have also shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC. Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome before and after surgical removal of lesion (adenoma or carcinoma) and whether the community changes at all.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) samples removal of lesion (adenoma or carcinoma).  Our analysis included both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest to build models to classify either lesion and normal samples or initial and follow up samples and then subsequently observe how these models, as well as specific OTUs within them, performed pre and post lesion removal.  We also investigated the models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normalsamples but also initial and follow up.

>>>>

**Results:** The bacterial microbiome changed more in response to lesion removal in carcinoma cases compared to adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between the pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial versus follow up samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.  Follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.  The initial sample model had a decrease in positive probability for the follow up samples tobe an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g. *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).

**Conclusions:** Our data supports the hypothesis that there are changes in the bacterial microbiome following colorectal cancer lesion removal.  Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.  Changes to commensal bacteria were some of the most important variables for model classification, suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].

Over the last few years there has been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have also shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.  Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion (adenoma or carcinoma).  We analyzed changes in alpha and beta diversity as well as specific Operational Taxonomic Units (OTUs) relative abundance.  We utilized Random Forest to build two models.  The first model was built to classify either lesion and normal samples while the second was built to classify initial and follow up samples.  Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.  We then investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.
